\bf A note on the instability and degeneracy of deep learning models

### 2017-09-27 writing/paper-note/docs/note.tex
some changes in the note from nordman

To define or measure instability in FSFS models, it is useful to
consider the behavior of a data model sequence \(P_{\theta_N}\). A
relevant quantity to this end is a (scaled) extremal log-probability
ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eq:elpr}
\end{align}
The main idea is that, in formulating FSFS models for potentially
increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
the ratio (\eqref{eq:elpr}) should remain bounded to ensure model
stability, requiring that the largest probability possible under
\(P_{\theta_N}\) maintain a fixed order of magnitude relative to the
smallest probability allowed under the same model. Specifically, the log
of the ratio should grow at mostly linearly with the sample size \(N\).
This leads to the following definition.
\BeginKnitrBlock{definition}[S-unstable FSFS]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FSFS) \fi{} }Let
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
model parameters where the size of the model \(q(N)\) is a function of
the number of random variables \(N\). A FSFS model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
variables increase (\(N \rightarrow \infty\)),
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a model is S-unstable if,
given any \(C > 0\), there exists an integer \(N_C > 0\) so that
\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
fails to be S-unstable.

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
non-exponential family models (e.g.~RBM and DBM models in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an
increasing number of parameters. While this definition differs in form
and scope from the original, it does match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models}) considered there.

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab-elpr}{}{\label{prp:instab-elpr}}Let
\(\text{ELPR}(\boldsymbol \theta_N)\) be as in (\eqref{eq:elpr}) for an
integer \(N \ge 1\). For a given \(C>0\), if
\[\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,\] then
\[\Delta_N(\boldsymbol \theta_N) > C.\]
\EndKnitrBlock{proposition}

Again, if the probability ratio (\eqref{eq:elpr}) is too large, then the
FSFS model will exhibit large changes in probability for very small
differences in the data configuration, which exemplifies the intuitive
notation of instability.

Additionally, S-unstable FSFS model sequences are connected to
degenerate models, where model \emph{degeneracy} typically entails
placing all probability on a small portion of the sample space. For
perspective, note that differing sizes of
\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in (\eqref{eq:elpr}) may
induce a spectrum of levels of ``stability'' and Proposition
\ref{prp:instab-elpr} indicates increasing sensitivity of model
probabilities (i.e., for one component changes in outcomes) as
(\eqref{eq:elpr}) increases. Furthermore, as the instability measure
(\eqref{eq:elpr}) grows, FSFS model sequences are guaranteed to slide into
full degeneracy as Proposition \ref{prp:degenFSFS} shows. Define a
\(\epsilon\)-modal set \[
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
\] of possible outcomes, for a given \(0 < \epsilon < 1\).

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:degenFSFS}{}{\label{prp:degenFSFS}}For an unstable
FSFS model in Definition \ref{def:instabFSFS}, and for any given
\(0 < \epsilon < 1\), \[
P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
\]
\EndKnitrBlock{proposition}

@@

\textbf{Proof of Proposition \ref{prp:degenFSFS}.} where
\(|\mathcal{X}|<\infty\) holds in the FSFS model. We may suppose
\(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
because otherwise the model is trivially degenerate for all
\(N \geq 1\). Fix \(0 < \epsilon < 1\). Then,
\(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
If
\(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\),
then by definition
\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\)
holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
\end{align*}
From the lower bound on
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
upper bound on
\(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it
follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
\end{align*}
as \(N \rightarrow \infty\) by the definition of an unstable FSFS model
(cf.~Definition \ref{def:instabFSFS}). Consequently,
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)

>>>>

To define or measure instability in FSFS models, it is useful to
consider the behavior of a data model sequence \(P_{\theta_N}\). A
relevant quantity to this end is a (scaled) extremal log-probability
ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eq:elpr}
\end{align}
The main idea is that, in formulating FSFS models for potentially
increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
the ratio \eqref{eq:elpr} should remain bounded to ensure model stability,
requiring that the largest probability possible under \(P_{\theta_N}\)
maintain a fixed order of magnitude relative to the smallest probability
allowed under the same model. Specifically, the log of the ratio should
grow at mostly linearly with the sample size \(N\). This leads to the
following definition.
\BeginKnitrBlock{definition}[S-unstable FSFS]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FSFS) \fi{} }Let
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
model parameters where the size of the model \(q(N)\) is a function of
the number of random variables \(N\). A FSFS model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
variables increase (\(N \rightarrow \infty\)),
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a model is S-unstable if,
given any \(C > 0\), there exists an integer \(N_C > 0\) so that
\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
fails to be S-unstable.

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
non-exponential family models (e.g., RBM and DBM models in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an
increasing number of parameters. While this definition differs in form
and scope from the original, it does match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models}) considered there.

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab-elpr}{}{\label{prp:instab-elpr}}Let
\(\text{ELPR}(\boldsymbol \theta_N)\) be as in \eqref{eq:elpr} for an
integer \(N \ge 1\). For a given \(C>0\), if
\[\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,\] then
\[\Delta_N(\boldsymbol \theta_N) > C.\]
\EndKnitrBlock{proposition}

Again, if the probability ratio \eqref{eq:elpr} is too large, then the
FSFS model will exhibit large changes in probability for very small
differences in the data configuration, which exemplifies the intuitive
notation of instability.

Additionally, S-unstable FSFS model sequences are connected to
degenerate models, where model \emph{degeneracy} typically entails
placing all probability on a small portion of the sample space. For
perspective, note that differing sizes of
\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in \eqref{eq:elpr} may
induce a spectrum of levels of ``stability'' and Proposition
\ref{prp:instab-elpr} indicates increasing sensitivity of model
probabilities (i.e., for one component changes in outcomes) as
\eqref{eq:elpr} increases. Furthermore, as the instability measure
\eqref{eq:elpr} grows, FSFS model sequences are guaranteed to slide into
full degeneracy as Proposition \ref{prp:degenFSFS} shows. Define a
\(\epsilon\)-modal set \[
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
\] of possible outcomes, for a given \(0 < \epsilon < 1\).

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:degenFSFS}{}{\label{prp:degenFSFS}}For an unstable
FSFS model in Definition \ref{def:instabFSFS}, and for any given
\(0 < \epsilon < 1\), \[
P_{\boldsymbol \theta_N}\left((X_1, \dots, X_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
\]
\EndKnitrBlock{proposition}

@@

\textbf{Proof of Proposition \ref{prp:degenFSFS}.} Where
\(|\mathcal{X}|<\infty\) holds in the FSFS model, we may suppose
\(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
because otherwise the model is trivially degenerate for all
\(N \geq 1\). Fix \(0 < \epsilon < 1\). Then,
\(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
If
\(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\),
then by definition
\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\)
holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
\end{align*}
From the lower bound on
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
upper bound on
\(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it
follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
\end{align*}
as \(N \rightarrow \infty\) by the definition of an S-unstable FSFS
model (cf.~Definition \ref{def:instabFSFS}). Consequently,
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
